Modeling the effect of aneuploidy on cancer evolution

### 2023-01-01 EvolutionaryRescue1.tex
Add files via upload

@@

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
\end{align}
\subsection{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{P_est.pdf}
\caption{}
\label{P_est}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{DeleteriousPlot.pdf}
\caption{}
\label{DeleteriousPlot}
\end{figure}
\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
\end{equation}
and using the following Taylor series expansion:
\begin{align*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
\end{align*}
we obtain:
\begin{align*}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&\approx\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}
\end{align*}
In the last line of the above we have used the following inequalities:
\begin{align*}
\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
\end{align*}
As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
\begin{align*}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
&=-\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_w-\mu_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
&=-\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
\end{align*}
where in the last line we have used the notations defined in \eqref{notationalconv}.

Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align*}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right].
\end{align*}

\begin{equation}
p_{est}\sim\left\{
  \begin{array}{@{}ll@{}}
  1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
  \end{array}\right.
\end{equation}
\section{Logistic}
We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
\begin{align*}
&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
\end{align*}
where $C_1, C_2>0$ are constants.
\section{Standing genetic variation}
Until now we have assumed that the initial population of cells consisted entirely of wildtype cells. We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy. The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
\begin{equation*}
p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
\end{equation*}
The total probability of evolutionary rescue is given by:
\begin{align}\nonumber
p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
\end{align}
The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
\begin{align*}
F\left(f\right)=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
\end{align*}
We let $F=\frac{1}{2}$ and we obtain:
\begin{equation*}
f^*\approx\frac{p_w}{p_w+p_a},
\end{equation*}
wher we have used the expansion $\e^x\approx 1+x$.
We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.

>>>>

Chromosomal instability (CIN) is the process during mitosis in which cells suffer from chromosome segration defects, which leads to a state called aneuploidy where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.

@@

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
\end{align}
\subsection{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{AneuploidyDeleteriousApprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{P_est.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
\label{P_est}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{P_est_divergence.pdf}
\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
\label{P_est}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{DeleteriousTauLeapPlot.pdf}
\caption{}
\label{DeleteriousPlot}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
\end{equation}
and using the following Taylor series expansion:
\begin{align*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
\end{align*}
we obtain:
\begin{align*}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)
\end{align*}
%In the last line of the above we have used the following inequalities:
%\begin{align*}
%\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
%\end{align*}
As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
\begin{align*}
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
\end{align*}
where in the last line we have used the notations defined in \eqref{notationalconv}.

Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align*}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right].
\end{align*}

\begin{equation}
p_{est}\sim\left\{
  \begin{array}{@{}ll@{}}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
  \end{array}\right.
\end{equation}
\section{Logistic}
We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
\begin{align*}
&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
\end{align*}
where $C_1, C_2>0$ are constants.
\section{Standing genetic variation}
Until now we have assumed that the initial population of cells consisted entirely of wildtype cells. We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy. The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
\begin{equation*}
p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
\end{equation*}
The total probability of evolutionary rescue is given by:
\begin{align}\nonumber
p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
\end{align}
The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
\begin{align*}
F\left(f\right)=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
\end{align*}
We let $F=\frac{1}{2}$ and we obtain:
\begin{equation*}
f^*\approx\frac{p_w}{p_w+p_a},
\end{equation*}
wher we have used the expansion $\e^x\approx 1+x$.
We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.
